A Linear Technique to Understand Non-Normal Turbulence Applied to a Magnetized Plasma

### 2014-03-04 Referee_Response1/referee_response.tex

{\bf Response to Referees \\
 \\} 
Brett Friedman and Troy Carter
%Updated: \today \\

\hrulefill

\section{Referee \#1}

\emph{This is an interesting paper, bringing non-modal analysis to bear upon a problem of
plasma turbulence. Non-modal analysis has been used very effectively in hydrodynamics.
Its use in plasma physics is a relatively recent trend, and holds the promise of providing
new perspectives on old problems. This paper, which includes theory as well as a strong
experimental connection, is consistent with the criteria for publication in PRL. Having
said that, we would like to request the authors to address the following questions:} \\

@@

\emph{1) After advocating the use of non-modal structures rather than linear eigenmodes for the
entirety of the paper, the author use a time-scale derived from linear eigenmodes (see
page 4, paragraph 2). Furthermore, they use the frequency for the n=1 mode instead of
that for the n=0 mode(s), with no apparent justification (other than "nonlinear coupling”).
It would seem much more consistent to use the time-scale for the nonlinear instability
they talk about earlier (although this would mean perhaps a loss of predictability that they
are very keen on). To us, this appears to be the weakest link in the paper. Perhaps, if they
present this paper as a study of the turbulence properties instead, they could give
information on how the curves in Figure 3 change with the choice of the linear time-
scale. It would seem that this would be much more useful to the reader in assessing the
power of the non-modal analysis, rather than presenting just two choices, and claiming
predictability. } \\

Response: First we would like to thank the referee for the careful reading of the paper and the nice review. As for the first comment, we believe that this is a valid criticism. 
We thought a lot about this before submitting the first manuscript and tried a few different linear time scales. The one we chose worked best, but it is difficult to justify it otherwise.
We agree that this was the weakest link of the paper. We also thought that the suggestion of presenting several choices of linear time scales was a good one, so this is what we've done.
We've presented four different choices of linear time scale and compared the results to each other. We have also softened our language on predictive capability and presented our results
as a way of understanding the turbulent growth rate through linear non-modal calculations with the ultimate goal of achieving prediction. Overall, we have modified the manuscript quite
a bit to make these changes, including changing the last two figures. \\

@@

\emph{2) What closure assumption is used to write equation (5)?} \\

Response: If by closure, you mean an approximation of the
nonlinearity, then we do not use any closure to write equation 5.  It is an exact equation that can be derived from the model equations~\cite{friedman2012b,friedman2013}.
We have also now written it differently to make it simpler and more transparent we hope. \\

@@

\emph{4) We think a log scale should be used for Figure 2. On the scale used in the paper, it is
impossible to determine if the growth rate at short times is somewhat similar to $G_{max} (t)$
or quite different. If similar, the potential for formulating fully analytic models is more
promising.} \\

Response: We have changed Figure 2 so that we do not show $G_{max}(t)$ any longer. The growth rate of the ensemble of curves is not close to the growth rate of $G_{max}(t)$. $G_{max}(t)$  grows so quickly,
and it isn't clear how one would use it in a procedure like ours because the points along it require linear evolution from very specific initial conditions, and every point comes from a different initial condition. \\

\emph{5) Minor point: Figure 3 is confusing simply because the important curves are not
sufficiently well-delineated. For example, the turbulent growth rate spectrum should be
bold (or something equivalent) to distinguish it from the others.} \\

Response: This figure is also much different now, but we have tried to make all of the curves distinguishable by using different colors along with symbols so that they are even distinguishable in black and white.

@@

\beq
\diff{\mathbf{u}}{t} = \mathbf{A u} \nonumber
\eeq

\emph{I would expect the prediction of the short time behavior of some norm of $\mathbf{u}$ through the mathematical analysis of $\mathbf{A}$
by methods other than the standard eigenmode decomposition, e.g., by methods based on the $\epsilon$-pseudospectrum and/or numerical range.}

\emph{If I understand correctly, the method presented here starts by dropping the nonlinear terms in
the equations of motion of the reduced Braginskii 2-fluid model. The resulting linear system is
numerically simulated up to a time $\tau_{nl}$ at which the system is then ‘re-randomized’ — whatever
it means. A growth rate $\gamma_{TR}$ defined and computed using data between $t = 0$ and $t = \tau_{nl}$ then
constitutes the prediction by this method. So it seems that the ‘non-modal method’ described in
this manuscript is basically a numerical simulation of the linearized equations of motion. While
prediction of turbulent properties by numerical simulations can be useful, it probably has less
impact than a theoretical prediction (which is admittedly much more difficult).} \\

Response: First, we would like to thank this referee for the careful reading and insights into the paper. As for the question over 'non-modal analysis', 
we contend that our method (especially with revisions in the new manuscript) is a form of 'non-modal analysis'. That is, we use only linear calculations to find growth rates without using normal modes (eigenvalues and eigenvectors). While most non-modal analysis in the literature seems to center around the calculation of the $epsilon$-pseudospectrum, much also includes the calculation of $||e^{\mathbf{A} t}||$ and $||e^{\mathbf{A} t} \mathbf{u}(0)||$ with various optimal $\mathbf{u}(0)$ vectors, i.e. to get the numerical and numerical abscissa. More importantly, however, it would be helpful to have a more theoretical prediction that doesn't require simulation. In that regard, we have come across a method~\cite{camargo1998} that can replicate our numerical simulations with a simple calculation that looks much more like a typical non-modal calculation. See Eqs. 9 and 10 and Figure 2. With this new change, we believe it is very clear that what we are doing is non-modal analysis and that we are not simply using linear simulations. \\

@@

\emph{2) For the procedure described above, there are several things I do not fully understand. First of
all, if $\gamma_{TR}$ is only computed using data for $0 < t < \tau_{nl}$, how does the ‘re-randomization’ at $t = \tau_{nl}$
come into play? I am probably missing something here.} \\

Response: We agree that this was a confusing way to word our
procedure. We therefore have removed this statement. At the beginning of the manuscript we offer this description of our model:
`` We model the turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, (2) linear transient growth deterministically amplifies the turbulent energy (or
decrease it in wavenumber ranges where linear damping dominates), (3) nonlinear transfer sets in at a specified timescale, terminating the transient growth process and re-randomizing the turbulent state (at which point the cycle repeats).`` This is not the procedure we use to do the actual calculation, but just our simplified description of the turbulent model.

As for the actual procedure, we state: ``In practice, we implement this model by starting with an ensemble of random initial conditions, which we evolve linearly for a time $\tau_{nl}$, 
and then take the time and ensemble averaged growth rate of these curves.''

Also, we use the non-modal calculation discussed above, added more equations, and replaced Figure 2 to make our procedure more clear. Please read the manuscript and see if this is now clear.
\\

@@

\beq
\gamma(m,n) = \frac{Q(m,n) + D(m,n)}{E(m,n)} \nonumber
\eeq
\emph{whereas the evolution of the energy for mode $(m, n)$ is given by}

\beq
\diff{E(m,n)}{t} = Q(m,n) + D(m,n) + \sum_{m',n'} T(m,m',n,n') \nonumber
\eeq
\emph{What is the reason of ignoring $T$ in the definition of $\gamma$? Similarly, what is the meaning/significance
of plotting $E(m, n, t)$ with nonlinearities turned off?  $\sum_{m',n'} T(m,m',n,n')$ is non-zero and contributes to 
$\diff{E(m, n)}{t}$ even though $\sum_{m,m',n,n'} T(m,m',n,n') = 0$. Would it be more informative
to compare $\gamma_{TR}$ to $\diff{E(m, n)}{t}$ instead of the $\gamma_e (m, n)$ defined in the manuscript (which ignores
contribution from $T$ )?} \\

Response: Note that we have changed our notation in the text, and added some more equations and discussions of this so that this issue should now be clear. But we will explain here anyway.

The reason why $T$ is missing from the definition of $\gamma$ is because $\diff{E(m,n)}{t} = Q(m,n) + D(m,n) + \sum_{m',n'} T(m,m',n,n')$ is virtually zero when the turbulence is in steady state (by definition of steady state). It may not be zero at a given time, but it is zero on average, and thus not suitable for use in defining the growth rate spectrum, which is a time-averaged property (see Figure 1a for the average and the spread in the growth rate). If we were to include $T$ in the definition, the turbulent growth rate curves in Figure 1a would be zero with some finite spread.

More informatively, the growth rate we use is a measure, by wavenumber, of the rate at which the fluctuations take energy from the equilibrium gradients (or return energy if the growth rate is negative). This comes purely from the linear terms. The nonlinear terms transfer energy between different fluctuation 'modes', but do not affect the overall energy budget.
So the picture is that certain modes or structures (those with
positive $\gamma$) inject energy from the equilibrium gradients into
the fluctuations through the linear terms. Those modes then
non-linearly transfer energy to other modes in an overall conservative
way (like a cascade). When energy is nonlinearly transferred into
modes with negative growth rate, that energy is dissipated. In the
case of normal linear operators, the unstable linear eigenvectors
inject energy, the nonlinearities transfer that energy to stable
linear eigenvectors which then dissipate the energy. For normal
operators, one would expect this growth rate to be pretty similar to
the fastest growing eigenmode growth rate, though it could be
smaller. But it certainly cannot be larger. We have added and modified
quite a bit of text and added a reference~\cite{terry2006b}, which
uses this growth rate definition.

@@

Finally, the reason we plot $E(m,n,t)$ with the nonlinearities turned off is to show that the linear terms alone cause (transient) growth of the fluctuations despite the fact that the $n=0$ mode numbers contain no unstable eigenmodes. It illustrates that the turbulent structures inject energy into the fluctuations in the short term. \\

\emph{4) I also feel that the clarity of the present manuscript suffers because of the page limitation of
PRL, a lot of the symbols are not defined (e.g. $v_E$, $N_r$ ) making it difficult for the general audience.
Together with the concerns raised above, I am afraid I cannot recommend the publication of the
present manuscript in the PRL.} \\

Response: We have now defined all symbols and removed unnecessary parts of the paper and replaced them with better text. We hope the major changes have made the manuscript significantly better and easier to understand.

### 2014-03-04 Transient_growth_prl.tex
March 4

\begin{abstract}
In nonlinear dynamical systems with highly nonorthogonal linear eigenvectors, linear non-modal analysis is more useful than normal mode analysis in predicting turbulent properties. 
However, the non-trivial time evolution of non-modal structures makes quantitative understanding and prediction difficult. 
We present a technique to overcome this difficulty by modelling the effect that the advective nonlinearities have on spatial turbulent structures. The nonlinearities are taken as a periodic randomizing force with
time scale consistent with critical balance arguments. We apply this technique to a model of drift wave turbulence in the Large Plasma Device (LAPD) 
[W. Gekelman \emph{et al.}, Rev. Sci. Inst. {\bf 62}, 2875 (1991)], where non-modal effects dominate the turbulence.
We compare the resulting growth rate spectra to that obtained from a nonlinear simulation, showing good qualitative agreement.
\end{abstract}

@@

Normal mode analysis -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has been used to solve many problems over the years.
Despite its wide-ranging success, it has failed in important instances, particularly in predicting the onset of subcritical turbulence in hydrodynamic flows. 
The reason for this failure was explained in the early 1990's when Trefethen and others attributed the pitfalls of normal mode analysis to the non-normality of linear operators of
dynamical systems~\cite{trefethen1993,schmid2007}. A non-normal operator has 
eigenvectors that are nonorthogonal to one another. One consequence of eigenvector nonorthogonality is that even when all eigenvectors decay exponentially under linear evolution, 
superpositions of eigenvectors can grow, albeit transiently.
In other words, certain fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
When combined with nonlinear effects, this allows for sustained subcritical turbulence.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under the 
action of the linear operator. Transient growth behavior, which can dominate turbulent evolution, can be discovered only through non-modal calculations.

Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
still heavily relies on normal mode analysis to inform turbulent predictions and observations, with a few notable exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing an approach to understand turbulent properties using only non-modal linear calculations with the goal of 
ultimately making quantitative predictions. 
Our approach is to evolve the linear solution from an ensemble of random initial conditions 
and calculate the average growth rate of the solutions over a specified timescale.  Optimally this time scale would be the
nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic \emph{linear} time. Since there is no obvious single linear time scale, we test several and compare the results to determine which works best.
The procedure ultimately produces growth rate spectra that can be used to predict turbulent properties such as saturation levels and transport rates through mixing length arguments.
While the concepts behind this technique are general enough to be applied to many nonlinear dynamical systems, the details vary for each case, 
so we restrict our treatment to one particular turbulence model. For this model, the technique qualitatively reproduces the turbulent growth rate spectrum of the direct 
nonlinear simulation fairly well, especially in comparison to the linear eigenmode spectrum.

@@

The nonlinear simulation reveals a fascinating property of the turbulence -- 
it is dominated by a nonlinear instability process despite being linearly unstable to drift waves~\cite{friedman2012b,friedman2013}.
The nonlinear instability, which was discovered by Drake et al.~\cite{drake1995}, works as follows: 
magnetic-field-aligned ($k_\para=0$) convective filaments transport density across the equilibrium density gradient, setting up $k_\para=0$ density filaments. 
These filaments are unstable to secondary drift waves, which grow on the periphery of the filaments. 
These drift waves, which have finite $k_\para$, nonlinearly couple to one another and reinforce the original convective filaments.

@@

From, Eqs.~\ref{dEdt_def} and~\ref{steady_state}, it follows that $ \gamma(m,n) = \gamma_l(m,n) + \gamma_{nl}(m,n) = 0$.
The rate of energy injection $\gamma_l(m,n)$ from the equilibrium gradient into the fluctuations is a quantity of great interest~\cite{friedman2012b,terry2006b}. 
When positive for some wavenumber, turbulence can be sustained. Additionally, it may be used in the mixing length formula $\gamma/k_\perp^2$ to predict the turbulent saturation level, 
and it is proportional to the transport rate~\cite{terry2006b}. 
$\gamma_l(m,n)$ may be calculated from the spatial structures of the plasma state variables, so it is always well-defined, even in a turbulent plasma. We plot
$\gamma_l(m,n)$ in Fig.~\ref{n0_n1_growth} a) for both the linear and steady state turbulent stages of our nonlinear simulation. The linear stage, which occurs when fluctuations
are small and exponentially growing has a time-independent $\gamma_l(m,n)$ equal to $\gamma_s(m,n)$ -- the ``spectral'' growth rate of the fastest growing eigenmode at each $m,n$.
During the turbulent stage, $\gamma_l(m,n)$ is time-dependent (indicated by the $1 \sigma$ spread), and generally much different than $\gamma_s(m,n)$. Significantly,
the turbulent $\gamma_l$ is positive at $n=0$ and low $m$ despite the fact that all linear eigenmodes have $\gamma_s < 0$ for $n=0$. This is a manifestation
of non-normality, for in normal systems, $\gamma_l(m,n) \le \gamma_s(m,n)$. Physically, it is the manifestation of the nonlinear
instability, specifically the linear part of the mechanism in which convective filaments drive density filaments from the equilibrium density gradient.

@@

Since transient growth is a purely linear phenomenon, it has been a goal of researchers to understand and predict the onset of subcritical turbulence using only linear, non-modal calculations. 
In our system, the turbulence is not subcritical in the traditional sense, but it has a subcritical component because $\gamma_l(m<50,n=0) > 0$, yet $\gamma_s(m,n=0) < 0$. 
It is our goal, then, to use linear non-modal calculations to understand this behavior and to move toward predictive capability of $\gamma_l$.
To accomplish this, we make the anzatz that nonlinearities randomize the turbulent spatial structure at each wavenumber on a time scale of 
one eddy decorrelation time, while the linearities evolve the spatial structures deterministically. From this we can calculate a non-modal $\gamma_{\rm{n-m}}$ spectrum,
which is our prediction of the turbulent $\gamma_l$ spectrum. To illustrate,
we begin by taking Eqs.~\ref{ni_eq}-\ref{te_eq} and Fourier decomposing in the azimuthal and axial directions.  Then, we discritize in the radial direction and
approximate radial derivatives with finite differences.  The resulting system of equations may be written in matrix form:

@@

\beq
\label{lin_eq_A}
\diff{\mathbf{u}}{t} = \mathbf{A} \mathbf{u},  \quad \rm{where} \ \mathbf{A} = \mathbf{M}^{\frac{1}{2}} \mathbf{B}^{-1} \mathbf{C} \mathbf{M}^{-\frac{1}{2}}.
\eeq
The solution of Eq.~\ref{lin_eq_A} is $\mathbf{u}(t) = e^{\mathbf{A} t} \mathbf{u}(0)$, which
depends on the initial condition $\mathbf{u}(0)$. For purposes of turbulent growth rate prediction, we are interested in
the behavior of $G(t) = E(t)/E(0) = ||\mathbf{u}(t)||^2/||\mathbf{u}(0)||^2$ and its rate of growth.

@@

It is common practice in normal mode analysis to look for the least stable eigenmode. 
For the non-modal case, it is common to study the properties of $G_{\rm{max}}(t) = ||e^{\mathbf{A} t}||$ because if this is greater than unity at any time, fluctuations may be amplified, 
leading to subcritical turbulence~\cite{trefethen2005,schmid2007}.
However, it can be misleading to study only $G_{\rm{max}}(t)$ when predicting specific properties of turbulence because
$G_{\rm{max}}(t)$ is only the upper envelope of all possible $G(t)$ curves. No one particular initial condition $\mathbf{u}(0)$ evolves along $G_{\rm{max}}(t)$. 
Furthermore, it isn't obvious what kind of spatial structure or structures will come to dominate a turbulent system.
In non-normal systems, unlike in normal systems, optimal structures don't amplify themselves, rather, they evolve while increasing the total fluctuating energy.

We contend that the key to understanding and predicting turbulent properties through non-modal analysis is 
to successfully model the effect that the nonlinearities have on the transient linear processes. 
To this effect, we note that the advective nonlinearity in Eq.~\ref{mat_eq} has the form of the state vector divided by a time $\tau_{nl} \sim (v_E k_\perp)^{-1}$. This nonlinear
time scale is generally associated with the eddy turnover or decorrelation time. We therefore present a heuristic model of the nonlinearities 
as a randomizing force that acts on this characteristic nonlinear time scale.
Specifically, we start with an ensemble of random initial conditions, evolve them linearly for a time $\tau_{nl}$, then take the time and ensemble averaged growth rate of these curves.
This procedure calculates the average growth rate of the turbulence as if the turbulence evolved linearly for a time $\tau_{nl}$, randomized, and then repeated. 
Now, this procedure does not require actual linear simulations from an ensemble of random initial conditions. To see this, recall that the time evolution of the energy from an initial condition is

@@

\beq
\label{E_t_ensemble_avg}
\left< E(t)/E(0) \right>_{{\rm{ens}}} = \frac{1}{4 N_r} {\rm{tr}} \{ e^{\mathbf{A} t} e^{\mathbf{A}^{\dagger}t} \}.
\eeq
We show the validity of this statistical averaging by plotting $1000$ ensemble curves in Fig.~\ref{m20n0_E_ev} along with their expected average from Eq.~\ref{E_t_ensemble_avg} and their actual average,
which line up perfectly.
One point to note is that in global equation sets like ours, where we discritize and use finite differences in the radial direction rather than a Fourier decomposition, randomizing $\mathbf{u}(0)$
amounts to setting the initial $k_r$ spectrum to a step function that goes to zero at the Nyquist wavenumber. This means that $\left< E(t)/E(0) \right>_{{\rm{ens}}}$ may depend on $N_r$, so we must be careful in
choosing $N_r$ for this analysis. We choose $N_r$ such that the grid spacing equals $\rho_s$, the general Nyquist wavenumber of drift wave simulations~\cite{scott1992}.

@@

Now there are several linear time scales that we may choose to test. We label these times in Fig.~\ref{m20n0_E_ev}
for the case of $n=0, m=20$. The first linear time is the linear eigenmode frequency, labeled
$\omega_s^{-1}$. However, $\omega_s = 0$ for $n=0$ linear eigenmodes, so we are forced to use $\omega_s$ for the fastest growing $n=1$ eigenmode at each $m$ to get a meaningful time scale.
Second is the parallel free-streaming time of the electrons $t_{\rm{st}} = L_\para/v_{t e}$ often cited in critical balance arguments~\cite{barnes2011}. Third is the time before modal effects
take over, which can be approximated as the time when $E(t)$ turns over. As seen in Fig.~\ref{n0_n1_growth}, $E(t)$ can be either a maximum or minimum before turning over. 
We label this time $E_{\rm{max}}$. 
Fourth, we use the steady-state condition $\gamma_l = - \gamma_{nl}$ and the approximation $\tau_{nl} \sim 1/|\gamma_{nl}|$ to get $\tau_{nl} = 1/|\gamma_{\rm{n-m}}|$. 
Inserting this into Eq.~\ref{gamma_l_calc} gives

@@

In summary, we present a procedure for calculating the turbulent growth rate spectrum using non-modal linear calculations. 
In the case of an LAPD experiment, this procedure captures the behavior of a nonlinear instability that dominates the dynamics of the turbulence.  In general,
non-modal analysis is difficult to quantify and make predictive, but using some simple nonlinear modelling, we have shown
that it may be possible. Future studies will attempt to test this procedure on other non-normal turbulence models and to see if it can predict critical parameters for subcritical turbulent onset.

>>>>

\begin{abstract}
In nonlinear dynamical systems with highly nonorthogonal linear eigenvectors, linear non-modal analysis is more useful than normal mode analysis in predicting turbulent properties. 
However, the non-trivial time evolution of non-modal structures makes quantitative understanding and prediction difficult. 
We present a technique to overcome this difficulty by modelling the effect that the advective nonlinearities have on spatial turbulent structures. 
The nonlinearities are taken as a periodic randomizing force with
time scale consistent with critical balance arguments. We apply this technique to a model of drift wave turbulence in the Large Plasma Device (LAPD) 
[W. Gekelman \emph{et al.}, Rev. Sci. Inst. {\bf 62}, 2875 (1991)], where non-modal effects dominate the turbulence.
We compare the resulting growth rate spectra to that obtained from a nonlinear simulation, showing good qualitative agreement, especially in comparison to the eigenmode growth rate spectra.
\end{abstract}

@@

Normal mode analysis -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has been used to solve many problems over the years.
Despite its wide-ranging success, it has failed in important instances, particularly in predicting the onset of subcritical turbulence in hydrodynamic flows. 
The reason for this failure was explained in the early 1990's when Trefethen and others attributed the pitfalls of normal mode analysis to the non-normality of linear operators of
dynamical systems~\cite{trefethen1993,schmid2007}. A non-normal operator has 
eigenvectors that are not orthogonal to one another. One consequence of eigenvector nonorthogonality is that even when all eigenvectors decay exponentially under linear evolution, 
superpositions of eigenvectors can grow, albeit transiently.
In other words, certain fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
When combined with nonlinear effects, this allows for sustained subcritical turbulence.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under  
action of the linear operator. Transient growth, which can dominate turbulent evolution, can be discovered only through non-modal calculations.

Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
generally relies on normal mode analysis to inform turbulent predictions and observations, with a few exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing an approach to understand turbulent properties using only non-modal linear calculations with the goal of 
ultimately making quantitative predictions.  Our approach is to calculate
an average growth rate of turbulent fluctuations due to linear
processes, specifically due to transient growth.  We model the
turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, (2) linear transient growth deterministically amplifies the turbulent energy (or
decrease it in wavenumber ranges where linear damping dominates), (3) nonlinear transfer sets in at a specified timescale, terminating the
transient growth process and re-randomizing the turbulent state (at which point the cycle repeats).    Optimally, the timescale for the final step would be the
nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic \emph{linear} time. Since there is no obvious single linear time scale, we test several and compare the results to determine which works best.
The procedure ultimately produces growth rate spectra that can be used
to predict turbulent properties such as saturation levels and transport rates through mixing length arguments.
While the concepts behind this technique are general enough to be applied to many nonlinear dynamical systems, the details vary for each case, 
so we restrict our treatment to one particular turbulence model. For
this model, the technique does reasonably well in reproducing the turbulent growth rate spectrum of the direct 
nonlinear simulation, especially in comparison to the linear eigenmode spectrum.

@@

The nonlinear simulation reveals a fascinating property of the turbulence -- 
it is dominated by a nonlinear instability process despite being linearly unstable to drift waves~\cite{friedman2012b,friedman2013}.
The nonlinear instability, which was discovered by Drake et al.~\cite{drake1995}, works as follows: 
magnetic-field-aligned ($k_\para=0$) convective filaments transport density across the equilibrium density gradient, setting up $k_\para=0$ density filaments. 
These filaments are unstable to secondary drift waves, which grow on the periphery of the filaments. 
These drift waves, which have finite $k_\para$, nonlinearly couple to one another and generate new convective filaments.

@@

From, Eqs.~\ref{dEdt_def} and~\ref{steady_state}, it follows that $ \gamma(m,n) = \gamma_l(m,n) + \gamma_{nl}(m,n) = 0$.
The rate of energy injection $\gamma_l(m,n)$ from the equilibrium gradient into the fluctuations is a quantity of great interest~\cite{friedman2012b,terry2006b}. 
When positive for some wavenumber, turbulence can be
sustained. Additionally, it may be used in a mixing length argument
to predict the turbulent saturation level, 
which can in turn be related to the rate of cross-field transport~\cite{terry2006b}. 
$\gamma_l(m,n)$ may be calculated from the spatial structures of the plasma state variables, so it is always well-defined, even in a turbulent plasma. We plot
$\gamma_l(m,n)$ in Fig.~\ref{n0_n1_growth} a) for both the linear and steady state turbulent stages of our nonlinear simulation. The linear stage, which occurs when fluctuations
are small and exponentially growing has a time-independent $\gamma_l(m,n)$ equal to $\gamma_s(m,n)$ -- the ``spectral'' growth rate of the fastest growing eigenmode at each $m,n$.
During the turbulent stage, $\gamma_l(m,n)$ is time-dependent (indicated by the $1 \sigma$ spread), and generally much different than $\gamma_s(m,n)$. Significantly,
the turbulent $\gamma_l$ is positive at $n=0$ and low $m$ despite the fact that all linear eigenmodes have $\gamma_s < 0$ for $n=0$. This is a manifestation
of non-normality, for in normal systems, $\gamma_l(m,n) \le \gamma_s(m,n)$. Physically, it is the manifestation of the nonlinear
instability, specifically the linear part of the mechanism in which convective filaments drive density filaments from the equilibrium density gradient.

@@

Since transient growth is a purely linear phenomenon, it has been a goal of researchers to understand and predict the onset of subcritical turbulence using only linear, non-modal calculations. 
In our system, the turbulence is not subcritical in the traditional sense, but it has a subcritical component because $\gamma_l(m<50,n=0) > 0$, yet $\gamma_s(m,n=0) < 0$. 
It is our goal, then, to use linear non-modal calculations to understand this behavior and to move toward predictive capability of $\gamma_l$.
To accomplish this, we make the anzatz that nonlinearities randomize the turbulent spatial structure at each wavenumber on a time scale of 
one eddy decorrelation time, while the linearities evolve the spatial structures deterministically. From this we can calculate a non-modal $\gamma_{\rm{n-m}}$ spectrum
which is our prediction of the turbulent $\gamma_l$ spectrum. To illustrate,
we begin by taking Eqs.~\ref{ni_eq}-\ref{te_eq} and Fourier decomposing in the azimuthal and axial directions.  Then, we discritize in the radial direction and
approximate radial derivatives with finite differences.  The resulting system of equations may be written in matrix form:

@@

\beq
\label{lin_eq_A}
\diff{\mathbf{u}}{t} = \mathbf{A} \mathbf{u},  \quad \rm{where} \ \mathbf{A} = \mathbf{M}^{\frac{1}{2}} \mathbf{B}^{-1} \mathbf{C} \mathbf{M}^{-\frac{1}{2}}.
\eeq
The solution of Eq.~\ref{lin_eq_A} is $\mathbf{u}(t) = e^{\mathbf{A} t} \mathbf{u}(0)$, which
depends on the initial condition $\mathbf{u}(0)$. For purposes of turbulent growth rate prediction, we are interested in
the behavior of $G(t) = E(t)/E(0) = ||\mathbf{u}(t)||^2/||\mathbf{u}(0)||^2$.

@@

It is common practice in normal mode analysis to look for the least stable eigenmode. 
For the non-modal case, it is common to study the properties of $G_{\rm{max}}(t) = ||e^{\mathbf{A} t}||$ because if this is greater than unity at any time, fluctuations may be amplified, 
leading to subcritical turbulence~\cite{trefethen2005,schmid2007}.
However, it can be misleading to study only $G_{\rm{max}}(t)$ when predicting specific properties of turbulence because
$G_{\rm{max}}(t)$ is only the upper envelope of all possible $G(t)$ curves. No one particular initial condition $\mathbf{u}(0)$ evolves along $G_{\rm{max}}(t)$. 
Furthermore, it isn't obvious what kind of spatial structures will come to dominate a turbulent system.
In non-normal systems, unlike in normal systems, optimal structures don't amplify themselves, rather, they evolve while increasing the total fluctuating energy.

We contend that the key to understanding and predicting turbulent properties through non-modal analysis is 
to successfully model the effect that the nonlinearities have on the transient linear processes. 
To this effect, we note that the advective nonlinearity in Eq.~\ref{mat_eq} has the form of the state vector divided by a time $\tau_{nl} \sim (v_E k_\perp)^{-1}$. This nonlinear
time scale is generally associated with the eddy turnover or decorrelation time. We therefore present a heuristic model of the nonlinearities 
as a randomizing force that acts on this characteristic nonlinear time scale. Again, this model is one in which 1) the turbulence begins as a random state, 
2) evolves linearly for a time $\tau_{nl}$, and 3) randomizes by nonlinear energy transfer, at which point the steps repeat.
In practice, we implement this model by starting with an ensemble of random initial conditions, which we evolve linearly for a time $\tau_{nl}$, 
and then take the time and ensemble averaged growth rate of these curves.
This procedure does not require actual linear simulations from an ensemble of random initial conditions. To see this, recall that the time evolution of the energy from an initial condition is

@@

\beq
\label{E_t_ensemble_avg}
\left< E(t)/E(0) \right>_{{\rm{ens}}} = \frac{1}{4 N_r} {\rm{tr}} \{ e^{\mathbf{A} t} e^{\mathbf{A}^{\dagger}t} \}.
\eeq
We show the validity of this statistical averaging by plotting an
ensemble of $1000$ curves generated with different random initial conditions in Fig.~\ref{m20n0_E_ev} along with their expected average from Eq.~\ref{E_t_ensemble_avg} and their actual average,
which agree well.
One point to note is that in global equation sets like ours -- where we discritize and use finite differences in the radial direction rather than a Fourier decomposition -- randomizing $\mathbf{u}(0)$
amounts to setting the initial $k_r$ spectrum to a step function that goes to zero at the Nyquist wavenumber. This means that $\left< E(t)/E(0) \right>_{{\rm{ens}}}$ may depend on $N_r$, so we must be careful in
choosing $N_r$ for this analysis. We choose $N_r$ such that the grid spacing equals $\rho_s$, the general Nyquist wavenumber of drift wave simulations~\cite{scott1992}.

@@

Now there are several linear time scales that we may choose to test. We label these times in Fig.~\ref{m20n0_E_ev}
for the case of $n=0, m=20$. The first linear time is the linear eigenmode frequency, labeled
$\omega_s^{-1}$. However, $\omega_s = 0$ for $n=0$ linear eigenmodes, so we are forced to use $\omega_s$ for the fastest growing $n=1$ eigenmode at each $m$ to get a meaningful time scale.
Second is the parallel free-streaming time of the electrons $t_{\rm{st}} = L_\para/v_{t e}$ often cited in critical balance arguments~\cite{barnes2011}. Third is the time before modal effects
take over, which can be approximated as the time when $E(t)$ turns over. As seen in Fig.~\ref{n0_n1_growth} b), $E(t)$ can be either a maximum or minimum before turning over. 
We label this time $E_{\rm{max}}$. 
Fourth, we use the steady-state condition $\gamma_l = - \gamma_{nl}$ and the approximation $\tau_{nl} \sim 1/|\gamma_{nl}|$ to get $\tau_{nl} = 1/|\gamma_{\rm{n-m}}|$. 
Inserting this into Eq.~\ref{gamma_l_calc} gives

@@

In summary, we present a procedure for calculating the turbulent growth rate spectrum using non-modal linear calculations. 
In the case of a simulation of an LAPD experiment, this procedure captures the behavior of a nonlinear instability that dominates the dynamics of the turbulence.  In general,
non-modal analysis is difficult to quantify and make predictive, but using some simple nonlinear modelling, we have shown
that it may be possible. Future studies will attempt to test this procedure on other turbulence models, and see if it can predict critical parameters for subcritical turbulent onset.
